Understanding fine temporal scale evolutionary dynamics with lineage metrics

### 2018-04-03 paper.tex

Yet, tracking the full details of a single lineage, much less a population of lineages, can be computationally expensive and will inevitably generate an unwieldy amount of data that can be challenging to visualize or interpret~\citep{mcphee_visualizing_2016}.
Summary statistics can help alleviate these issues by enabling the user to focus on aggregate trends across a population rather than needing to examine each individual's lineage. The question is how to effectively summarize %what is essentially 
a path through fitness space. One useful abstraction is to treat the path as a sequence of states. Here, we use phenotypes and genotypes as the states in the sequence, but we could just as easily use some other descriptor of the lineage's position in the fitness landscape at a given point in time.
With this abstraction in hand, a few metrics are easily formalized: the number of unique states, the number of transitions between states, and the amount of time spent in each state.
Additionally, we may care about how the transitions between states happened. What mutations led to them? Were those mutations beneficial, deleterious, or neutral at the time? These mutations are particularly notable because they did not simply appear briefly, but stood the test of time, leaving descendants in the final population.
%@CAO: Does the sentence I added above make sense?  My point being that these aren't just a random sampling of possible mutations that we are considering.  It's almost worth an extra sentence to point out that any deleterious mutations on the lineage are of special interest since they made it into the lineage despite being actively selected against.
% @AML: I think so. My interpretation: all mutations (beneficial, deleterious, neutral, etc., etc.) that appear on a lineage of an extant individual are particularly notable. 
%Here, we choose a subset of these metrics that we expect will be broadly useful to explore further.
Here, we explore a subset of these metrics that we expect will be broadly useful.

@@

%@AML:  It might be nice to wrap this section up with a little bit of a 'call to action'/'why should you care about this as a community'
%@ELD: I am in favor.
In addition to demonstrating a range of metrics that are immediately applicable and useful to digital evolution research, we intend for this work to begin a conversation within the %computational evolution
artificial life
community about how we quantify, interpret, and compare observed evolutionary histories. There has been extensive efforts to improve our ability to visualize both lineages and phylogenies [CITATIONS], which are %incredibly valuable for building intuitions and qualitatively understanding the dynamics embedded in a population's evolutionary history. However, we are unaware of efforts to formalize a suite of quantitative lineage and phylogeny-related metrics for computation evolution. Indeed, .... % people have done stuff with lineages
% - often a component of a larger experiment-specific analysis or limited to qualitative descriptions/visualizations
% - well-defined metrics not only provide valuable tools for teasing apart evolutionary dynamics, they also facilitate allow us to move away from exclusively qualitative descriptions/understandings of results ==> better quantitative understanding/statistics ==> better hypothesis testing
% - facilitate comparisons among systems. 
% - Something about activity metrics?

@@

Code for all of our metrics is open source and available in the Empirical library (https://github.com/devosoft/Empirical). Empirical is a C++ library built to facilitate writing efficient and easily sharable scientific software. To facilitate including useful features without excessive overhead, Empirical is header-only, meaning these metrics can be imported into an existing project with minimal effort.
% [QUESTION] @AML: Is it actually true that someone could include these metrics with minimal effort? How much of Empirical would it impose on the user? Just the systematics manager? 
%@ELD: They should just need the systematics manager. And they might need to hook up a couple of data collection signals.

\subsection{Lineage Metrics}
% Outlining some points to (maybe) make here; probably too bogged down in details; not even sure we want to talk about lineages as graphs:
% - Apply metrics in context of asexual populations
% - Establish a framework from which to think about metrics (not novel, but rarely formalized)
% 	 - We treat lineages as sequences of states where each state represents an individual and transitions between states represent parent-offspring relationships.
%      - More formally, lineages are directed acyclic graphs where internal nodes have in-degree = 1 (parent), out-degree = 1 (offspring). Original ancestor (in digital evolution, often seed of population) w/out-degree = 1. Extant individual w/in-degree = 1.
%      - Transitions specify relevant variation introduced into offspring (e.g. mutations)
%      - States specify relevant genotype and phenotype information (e.g. genome, location in space, behavior, etc) -- anything we might want to measure over a lineage
%      - Becomes more complicated with sexual populations, but we'll stick with asexual assumption for this. 
%  - Already, framework affords certain questions:
%     - How many individuals are along the lineage? 
%     - Total # and magnitude of mutations along lineage?
%     - Distribution of mutation types (ins vs. sub/beneficial vs. deleterious)?
%  - Further abstractions:
%      - Sequence of genotypes along lineage (where each genotype may represent multiple individuals)
%      - Sequence of phenotypes along lineage (where each phenotype may represent multiple individuals)
%      - Still further abstraction: sequence of a particular aspect of genotype (e.g. genetic marker)/phenotype (e.g. location) along lineage
%  - We discuss three classes of lineage metrics within this framework: lineage length, mutation accumulation, and phenotypic volatility. 
%  - For each, we discuss our intuition for what they can tell us (alone and in combination) about evolution. 
Each of the three lineage metrics we discuss --- lineage length, mutation accumulation, and phenotypic volatility --- reduces a lineage to a linear sequence of states where each state represents an individual or a continuous sequence of individuals that share some common genotypic or phenotypic characteristic of interest. While we limit our focus to three lineage metrics, this abstraction places lineages in a form suitable for a wide range of measurements, including the direct application of many data mining techniques designed to operate over sequences such as sequential pattern mining, trend analysis, \textit{et cetera} \citep{han2011data}.

Only lineages consisting of asexually reproducing ancestors where genetic material is exclusively vertically transmitted can be directly abstracted as a \textit{linear sequence} of states. 
Sexual reproduction (and any form of horizontal gene transfer) complicates matters significantly as such lineages are more appropriately represented by trees rooted at the extant organism, branching for each contributor of genetic material. 
For each metric we present here, we limit our discussion to asexual populations; however, we suggest two approaches for generalizing these lineage metrics to sexual populations: apply a lossy compression to reduce sexual lineages into linear sequences or extend our application of the metrics to operate over non-linear (tree) structured lineages. 
Sexual lineages can be compressed into linear sequences of states by modeling sexual reproduction events as asexual reproduction events, designating one parent to be a part of the lineage and considering the genetic contributions of other parents as sources of genetic variation (mutations). The primary downside to this approach is its lossy-ness (i.e. the fact that it discards potentially important parentage information). 
Alternatively, we can extend our metrics to operate over the more complex state sequences that constitute the lineages of sexually reproducing organisms. One such approach would be to consider all possible ancestor paths for an extant individual, calculating a given metric for each of them and then averaging the resulting values together. Assessing the efficacy of these and potentially other approaches would be a useful line of research to pursue in the future.

\subsubsection{Lineage Length}
% Define/explanation
% 	- Within context of framework, lineage length is simply measure of number of states in the graph/sequence.
%   - What this means will depend on what states represent?
%       - number of individuals along lineage 
%             - More interesting in experiments where generations are not dictated by system, but rather by life history strategies of organisms
%             - Give you a diagnostic for replication rate (r vs. k strategy)
%      - number of genotypes along lineage
%            - length = 1: no mutations
%            - length = number of individuals: mutation in every replication
%            - Intuitively, should be higher at high mutation rates, lower at low mutation rates. 
%      - number of phenotypes along lineage
%            - low: evidence for static environment
%            - high: might be indicative of a changing environment or ecological interactions
%            - if number of genotypes is high, but number of phenotypes low, probably in a neutral fitness landscape
%      - various subsets of genotype/phenotype
%           - location in space (environment type): useful for looking at lineage through heterogeneous landscape
%           - change in particular phenotypic trait
%           - change in genetic markers
%           - Distributions of above two lengths can tell you useful things about drift, selection, etc.

Lineage length describes the number of ancestral states traversed by a lineage. If a state represents a single individual along a lineage, lineage length is a measure of the number of generations on a lineage. This form of the lineage length metric is most useful in systems where generational turnover is not algorithmically determined but instead determined by the life history strategies of organisms. In these systems, lineage length can provide a proxy for replication rate where, for lineages that span equal lengths of time, longer lineages imply faster replication rates (\textit{e.g.} r-selected lineage) and shorter lineages imply slower replication rates (\textit{e.g.} K-selected lineage). 
%@ELD: I've always heard these called r- and k- selected rather than replicator, but I'm totally willing to believe these terms are used too.
%@AML: Yup. I have no idea if I've ever heard them referred to as 'r-replicator vs. k-replicator'. I'm going to get rid of the two parentheticals for now.
%@ELD: I like the idea of putting it in that context. I was just suggesting we replace "replicator" with "selected"
%@AML: Put the parentheticals back.

@@

In conjunction with collected fitness information, the class of a mutation (\textit{e.g.} beneficial, deleterious, or neutral) can also be tracked. Different evolutionary conditions are expected to cause different distributions of mutations along a lineage~\citep{barrick2013genome}; deviations revealed by measures of mutation accumulation can act as a barometer for unexpected evolutionary dynamics.  
% @AML: the below points likely have many things that could be cited to support. 
The number and magnitude of deleterious mutations along a lineage can both tell us something about the ruggedness of the fitness landscape, as well as how well a lineage was able to cross fitness valleys~\citep{covert_experiments_2013}. 
Similarly, an elevated measure of neutral mutations relative to beneficial or deleterious mutations can tell us that the fitness landscape has a large neutral area that the lineage is spending most of its time drifting around.

@@

Phenotypic volatility addresses the rate at which the lineage changes phenotype (although the same concept can be applied to other types of state). In systems with discrete/categorical phenotypes, this can be measured by summing the number of times the phenotype changes. A related but subtly different measurement in such systems is the number of unique phenotypes on a lineage. In most cases, these values will be very similar. A discrepancy between them, however, would be a clear signal of interesting dynamics. Specifically, it would suggest that the lineage was in a portion of the landscape where it could rapidly switch among different phenotypes. Such behavior could be indicative of some form of evolutionary bet-hedging~\citep{beaumont2009experimental}. % @AML: There are a whole lot of citations for bet-hedging.. Just picked a popular Nature article on experimental evolution of bet hedging. Alternatively, could go with older paper suggesting bet-hedging as a possible strategy that exists. 
%@ELD: This one sounds good. Normally I'd say cite both, but we probably don't have the space.

In systems with continuous-valued phenotypes, a subtly different approach is needed to measure phenotypic volatility, because there are no discrete state transitions. Instead, we can measure the overall variance in phenotype along a lineage. In some cases, it may be desirable to smooth out the noise inherent in a real-valued phenotype. We can do so by instead taking the variance of the rolling mean of fitness, to more closely approximate the idea of measuring phase transitions.
% SPJ: What is meant by 'rolling mean' ?
%@ELD: Would "moving average" be more familiar?

@@

Each of these metrics can be calculated for each member of the population at each time step. Doing so, however, would produce an amount of data so large that it would be difficult to make sense of. Instead, we need to come up with ways to generate useful summaries. There are two main approaches to doing so: 1) choose a small number of representative lineages from a given time point, 2) collect summary statistics about the distribution of metric values across the population.

A single representative lineage can be chosen by selecting the lineage of the dominant organism (either the most fit or the more numerous; here we use the most fit). In populations with a diversity of coexisting strategies, this approach can be uninformative as the dominant lineage is not necessarily representative of all successful lineages. An alternative in these scenarios is to filter out lineages that do not have offspring some predetermined number of generations later as such lineages were likely not representative of an important subset of the population. Still, any approach based on measuring only a subset of lineages can be challenging to interpret when the current dominant lineage (or lineages) is replaced with a different one; such changes can introduce a discontinuity if the value is being measured over time. If graceful responses to changes in which lineage is dominant are required, it can be advantageous to instead measure summary statistics (e.g. mean, variance, and range) across the entire population.

In scenarios where selective sweeps happen rapidly, the dominant lineage will likely be very similar to the average lineage in the population, as most of the population will be closely related. When the population contains more phylogenetic diversity, however, it is possible for the dominant lineage to differ from the mean. The nature of such differences is likely informative about the evolutionary dynamics occurring in the population. 
%Since we have yet to observe an instance of this phenomenon, we do not explore it further in this paper. 
% @AML: Not sure what the above sentence is saying.
%Nevertheless, it presents an argument for measuring the dominant lineage in addition to gathering summary statistics across the population. % @SPJ: I am not sure what argument this is referring to? The fact that there may or may not be a dominant lineage that differs from the mean?
%@ELD: The point was just that sometimes the dominant lineage is different from the mean and that probably matters but we're not going to talk about it. Since we're going to be short on space, I think we can just cut these lines.

@@

These metrics operate on entire phylogenies rather than single lineages within a population. Thus, in most cases, summarizing them across the population is not an issue as it is with lineage metrics. Because they factor in the entire population, phylogeny metrics can be more computationally expensive to calculate than metrics about a single lineage. On the other hand, because most lineages in the population share a substantial amount of history, phylogeny metrics can often be calculated more rapidly than lineage metrics can be calculated across an entire population. Note that phylogenies can be constructed with regard to any taxonomic level of organization, be it individual, genotype, phenotype, etc. Thus, when we refer generally to items in a phylogeny, we will use the term \textit{taxa}. % @SPJ: could reword to 'Thus, when we refer generally to items in a phylogeny, we will use the term \textit{taxa}.',  might be more clear?
%@ELD: sounds good!

A standard technique for saving memory and time when working with lineages in computational systems is to ``prune" them, removing any taxa that are not either currently alive or the ancestor of taxa that are alive. Since all of the phylogeny metrics we discuss here are borrowed from natural systems (where we do not have information about taxa without offspring), they all are designed to work on pruned phylogenies. Thus, for the remainder of this paper, we will assume we are only working with pruned phylogenies.

@@

An important distinction between phylogenies in natural versus computational systems to be aware of is that in natural systems phylogenies are generally inferred from extant taxa, whereas in computational systems they are directly measured. The inferred phylogenies used in natural systems do not contain internal nodes unless those nodes are branch points. They also do not contain history prior to the most recent common ancestor (MRCA) of all organisms in the extant population. For consistency, we will exclude pre-MRCA taxa from our analyses. However, we will not remove non-branching internal nodes, as these only serve to make our phylogenies more informative.

@@

The depth of the MRCA (i.e. how many steps from the beginning of time did it occur?) is a useful metric because of the relatively high ratio of informativeness to ease of calculation. A more recent MRCA implies more frequent selective sweeps and less long-term stable coexistence between clades. The depth of the MRCA is easily calculated because most implementations of phylogeny pruning require keeping track of it already. Measuring the frequency with which the MRCA changes (i.e. the number of coalescence events) can also be informative, as some conditions can inflate the length of the lineage relative to other conditions without actually increasing the frequency of selective sweeps. This scenario is particularly likely when the population size is changing over time. A downside to the depth of MRCA as a metric is that any population that does have a stable ecology will likely never see a change in the MRCA after the very beginning of evolution (which is informative as a way of detecting stably coexisting branches in the phylogeny). % @SPJ: change to 'The depth of the MRCA is easily calculated because most implementations of phylogeny pruning require ...' (i.e remove 'so' and 'most any')
% @SPJ: The above sentence might need use some support (why is it informative?)
%@ELD: Better?

@@

Measurements of phylogenetic richness quantify the total amount of evolutionary history that is contained in a set of taxa. The most traditional metric of phylogenetic richness is the one simply titled ``Phylogenetic Diversity", which is calculated as the number of the nodes in the minimum spanning tree from the MRCA to all extant taxa~\citep{faith_conservation_1992}. Another approach is to calculate the pairwise distances between all taxa and sum them~\citep{tucker_guide_2017}. A third approach is to calculate evolutionary distinctiveness, a measurement of a taxon's evolutionary uniqueness~\citep{isaac_mammals_2007}, and sum it across all extant taxa~\citep{tucker_guide_2017}.

@@

Measurements of phylogenetic divergence quantify how distinct the taxa in the population are from each other and are often the means of values that can be calculated for individual taxa. For instance, one option is to calculate the mean of the pairwise distances between all taxa in the population~\citep{webb_exploring_2000}. Similarly, phylogenetic divergence can also be calculated by calculating the evolutionary distinctiveness of each taxon in the population (as alluded to in the previous section) and taking the average of these values.

@@

\section{Test Problems}
\begin{figure}
\includegraphics[width=3.5in]{figs/landscapes.png}
\caption{The fitness landscapes used in this experiment: A) Himmelblau, B) Six-humped Camel Back, C) Shubert, and D) Composition Function 2. Interactive versions available at \url{https://emilydolson.github.io/fitness_landscape_visualizations}.}
\label{fig:landscapes}
\end{figure}

To understand the metrics defined above, the test problems used need to be well understood and studied. The benchmark functions from the GECCO Competition on Niching Methods meet both of these requirements and allow us to visualize the actual fitness landscape, due to the low dimensionality of the problems ~\citep{li_benchmark_2013}. For each problem, the X and Y coordinates offered by a given organism are translated by the function into a fitness value. We chose a diverse subset of these functions (Himmelblau, Shubert, Composition Function 2, and Six-Humped Camel Back) as our test problems in order to gain a broad understanding of our metrics. We used the implementations of these problems at https://github.com/mikeagn/CEC2013/ (C++ for fitness calculations during evolution, Python for post-hoc analysis). Figure \ref{fig:landscapes} shows the fitness landscapes defined by each of our four chosen test problems.

>>>>

Yet, tracking the full details of a single lineage, much less a population of lineages, can be computationally expensive and will inevitably generate an unwieldy amount of data that can be challenging to visualize or interpret~\citep{mcphee_visualizing_2016}.
Summary statistics can help alleviate these issues by enabling the user to focus on aggregate trends across a population rather than needing to examine each individual's lineage. The question is how to effectively summarize %what is essentially 
a path through fitness space. One useful abstraction is to treat the path as a sequence of states. Here, we use phenotypes and genotypes as the states in the sequence, but we could just as easily use some other descriptor of the lineage's position in the fitness landscape at a given point in time.
With this abstraction in hand, a few metrics are easily formalized: the number of unique states, the number of transitions between states, and the amount of time spent in each state.
Additionally, we may care about how the transitions between states happened. What mutations led to them? Were those mutations beneficial, deleterious, or neutral at the time? These mutations are particularly notable because they did not simply appear briefly, but stood the test of time, leaving descendants in the final population.
%@CAO: Does the sentence I added above make sense?  My point being that these aren't just a random sampling of possible mutations that we are considering.  It's almost worth an extra sentence to point out that any deleterious mutations on the lineage are of special interest since they made it into the lineage despite being actively selected against.
% @AML: I think so. My interpretation: all mutations (beneficial, deleterious, neutral, etc., etc.) that appear on a lineage of an extant individual are particularly notable. 
%Here, we choose a subset of these metrics that we expect will be broadly useful to explore further.
%@ELD: I agree. It's also nice to emphasize in light of deleterious mutations being the most interesting part of our results.
Here, we explore a subset of these metrics that we expect will be broadly useful.

@@

%@AML:  It might be nice to wrap this section up with a little bit of a 'call to action'/'why should you care about this as a community'
%@ELD: I am in favor.
In addition to demonstrating a range of metrics that are immediately applicable and useful to digital evolution research, we intend for this work to begin a conversation within the %computational evolution
artificial life
community about how we quantify, interpret, and compare observed evolutionary histories. There have been extensive efforts to improve our ability to visualize both lineages and phylogenies~\citep{standish_visualising_2002,burlacu_visualization_2013, mcphee_using_2016,mcphee_visualizing_2016, lalejini_evolutionary_2016}, which are %incredibly valuable 
indispensable 
for building intuitions and qualitatively understanding the dynamics embedded in a population's evolutionary history. However, we are unaware of efforts to formalize a suite of quantitative lineage and phylogeny-based metrics for computational evolution. %Indeed, .... % people have done stuff with lineages
% @CAO: For the moment I commented the "Indeed" out since I'm not sure where it's going; think the current version is a fine place to stop, though I also agree that some of the points below would be good to include if we can.
% - often a component of a larger experiment-specific analysis or limited to qualitative descriptions/visualizations
% - well-defined metrics not only provide valuable tools for teasing apart evolutionary dynamics, they also facilitate allow us to move away from exclusively qualitative descriptions/understandings of results ==> better quantitative understanding/statistics ==> better hypothesis testing
% - facilitate comparisons among systems. 
% - Something about activity metrics?

@@

Code for all of our metrics is open source and available in the Empirical library (\url{https://github.com/devosoft/Empirical}). Empirical is a C++ library built to facilitate writing efficient and easily sharable scientific software. To simplify including useful features with minimal overhead, Empirical is header-only, meaning these metrics can be imported into an existing project with limited effort.
% [QUESTION] @AML: Is it actually true that someone could include these metrics with minimal effort? How much of Empirical would it impose on the user? Just the systematics manager? 
%@ELD: They should just need the systematics manager. And they might need to hook up a couple of data collection signals.

\subsection{Lineage Metrics}
Each of the three lineage metrics that we discuss --- lineage length, mutation accumulation, and phenotypic volatility --- reduces a lineage to a linear sequence of states where each state represents an individual or sequence of individuals that share a common genotypic or phenotypic characteristic of interest; Figure \ref{fig:lineage_metrics_cartoon} is given as a toy example to help guide our discussion of these metrics. 
While we limit our focus to three lineage metrics, this abstraction places lineages in a form suitable for a wide range of measurements, including the direct application of many data mining techniques designed to operate over sequences such as sequential pattern mining, trend analysis, \textit{et cetera} \citep{han2011data}.

Only asexual lineages %consisting of asexually reproducing ancestors 
where genetic material is exclusively vertically transmitted can be directly abstracted as a \textit{linear sequence} of states.
Sexual reproduction (and any form of horizontal gene transfer) complicates matters significantly as such lineages are more appropriately represented by trees rooted at the extant organism, branching for each contributor of genetic material. 
% For the metrics we present here, we limit our discussion to asexual populations; however, we suggest two approaches for generalizing these lineage metrics to sexual populations: apply a lossy compression to reduce sexual lineages into linear sequences or extend our application of the metrics to operate over non-linear (tree) structured lineages. 
One possibility is to compress sexual lineages into linear sequences of states by modeling sexual reproduction events as asexual reproduction events, designating one parent to be a part of the lineage and considering the genetic contributions of other parents as sources of genetic variation (mutations). The primary downside to this approach is its lossy-ness (i.e. the fact that it discards potentially important parentage information). 
Alternatively, we can extend our metrics to operate over the more complex state sequences that constitute the lineages of sexually reproducing organisms. One such approach would be to consider all possible ancestor paths for an extant individual, calculating a given metric for each of them and then averaging the resulting values together. Another approach would be to divide an organism into its constituent parts that are inherited atomically (such as genes or instructions, depending on the representation); an organism would then be viewed as a collection of lineages rather than a single one. 
Assessing the efficacy of these and potentially other approaches would be a useful line of research to pursue in the future.

\begin{figure*}[t]
\includegraphics[width=\textwidth]{figs/lineage_metrics_cartoon.png}
\caption{\small Panels (A), (B), (C), and (D) each show different ways of representing a lineage. The lineage has accumulated three mutations: one reverse mutation and two substitutions. The phenotype of ancestors changed three times. In (A), each ancestral state along the lineage represents a single individual. Here, the lineage length is equal to the number of generations spanned by the lineage (eight). In (B), ancestral states represent the sequence of genotypes along the lineage. Here, the lineage length is equal to four. In (C) ancestral states represent the sequence of phenotypes along the lineage. Here, the lineage length is equal to the number of times a different phenotype is expressed along the lineage (three). In (D), ancestral states are given by a particular phenotypic characteristic. Here, lineage length is equal to two.}
\label{fig:lineage_metrics_cartoon}
\end{figure*}

\subsubsection{Lineage Length}
Lineage length describes the number of ancestral states traversed by a lineage. If a state represents a single individual along a lineage, lineage length is a measure of the number of generations on a lineage. This form of the lineage length metric is most useful in systems where generational turnover is not algorithmically determined, but instead determined by the life history strategies of organisms. In these systems, lineage length can provide a proxy for replication rate where, for lineages that span equal lengths of time, longer lineages imply faster replication rates (\textit{e.g.} r-selected lineage) and shorter lineages imply slower replication rates (\textit{e.g.} K-selected lineage). 
%@ELD: I've always heard these called r- and k- selected rather than replicator, but I'm totally willing to believe these terms are used too.
%@AML: Yup. I have no idea if I've ever heard them referred to as 'r-replicator vs. k-replicator'. I'm going to get rid of the two parentheticals for now.
%@ELD: I like the idea of putting it in that context. I was just suggesting we replace "replicator" with "selected"
%@AML: Put the parentheticals back.

@@

In conjunction with collected fitness information, the class of a mutation (\textit{e.g.} beneficial, deleterious, or neutral) can also be tracked. Different evolutionary conditions are expected to cause different distributions of mutations along a lineage~\citep{barrick2013genome}; deviations revealed by measures of mutation accumulation can act as a barometer for unexpected evolutionary dynamics.  
% @AML: the below points likely have many things that could be cited to support. 
The number and magnitude of deleterious mutations along a lineage can both tell us something about the ruggedness of the fitness landscape, as well as how well a lineage was able to cross fitness valleys~\citep{covert_experiments_2013}. 
Similarly, an elevated measure of neutral mutations relative to beneficial or deleterious mutations can tell us that the fitness landscape has a large neutral area that the lineage is spending most of its time drifting around. 
%@CAO There are various other things it could tell us, such as additional co-evolutionary pressures that are not reflected in the raw calculation of fitness.  Interestingly, using these metrics in combinaton with local fitness landscape analyses would do the trick.  But that's for another paper.  Indeed, I think it would be nice to put out a series of papers of useful analysis techniques and then do a master paper that reviews them all and demonstrates them in the context of an Avida-like system AND a very different digital evolution system (Markov Brains?)

@@

Phenotypic volatility addresses the rate at which phenotype changes as you move down a lineage (although the same concept can be applied to specific phenotypic traits or other types of state). In systems with discrete/categorical phenotypes, this can be measured by summing the number of times the phenotype changes. A related but subtly different measurement in such systems is the number of unique phenotypes on a lineage. In most cases, these values will be similar; a discrepancy
%between them, however, would be a clear signal of interesting dynamics. Specifically, it 
would suggest that the lineage was %in a portion of the landscape where it could rapidly switch among different 
cycling through a set of
phenotypes. Such behavior could, for example, be indicative of some form of evolutionary bet-hedging~\citep{beaumont2009experimental}. % @AML: There are a whole lot of citations for bet-hedging.. Just picked a popular Nature article on experimental evolution of bet hedging. Alternatively, could go with older paper suggesting bet-hedging as a possible strategy that exists. 
%@ELD: This one sounds good. Normally I'd say cite both, but we probably don't have the space.

In systems with continuous-valued phenotypes, a subtly different approach is needed to measure phenotypic volatility, because there are no discrete state transitions. Instead, we can measure the overall variance in phenotype along a lineage. In some cases, it may be desirable to smooth out the noise inherent in a real-valued phenotype. We can do so by instead taking the variance of the moving average of fitness, to more closely approximate the idea of measuring phase transitions.
% SPJ: What is meant by 'rolling mean' ?
%@ELD: Would "moving average" be more familiar? 
%@ELDL Also there turns out not to be much of a difference, so we could just not mention that approach.
% @SPJ: I don't think I have heard either of those terms. If we need space, it may be something to cut?
% @CAO: I suspect "moving average" is the more popular term, but I suspect people would figure out what we meant by either.
% @ELD: Okay, I changed it, but also this sentence is the first to go if we run into space issues.

@@

Each of these metrics can be calculated for each member of the population at each time step. Doing so, however, would produce an amount of data so large that it would be difficult to make sense of. Instead, we need to come up with ways to generate useful summaries. There are two main approaches to doing so: 1) choose a small number of representative lineages from a given time point, or 2) collect summary statistics about the distribution of metric values across the population.

A single lineage can be chosen by selecting the lineage of %the dominant
a representative
organism (either the most fit or the more numerous; here we use the most fit). In populations where diverse strategies coexist, this approach can be uninformative as any one lineage is unlikely to be representative of all successful lineages. One alternative %in these scenarios 
is to filter out lineages that do not have offspring some predetermined number of generations later as such lineages were likely not representative of an important subset of the population. Still, any approach based on measuring only a subset of lineages can be challenging to interpret when the current dominant lineage (or lineages) is replaced with a different one; such changes can introduce a discontinuity if the value is being measured over time. If graceful responses to changes in which lineage is dominant are required, it can be advantageous to instead measure summary statistics (e.g. mean, variance, and range) across the entire population.

In scenarios where selective sweeps happen rapidly, the dominant lineage will likely be similar to the average lineage,
%in the population,
as most of the population will be closely related. When the population contains more phylogenetic diversity, however, it is possible for the dominant lineage to differ from the mean. Of course, the nature of such differences is likely informative about the evolutionary dynamics occurring in the population. 
%Since we have yet to observe an instance of this phenomenon, we do not explore it further in this paper. 
% @AML: Not sure what the above sentence is saying.
%Nevertheless, it presents an argument for measuring the dominant lineage in addition to gathering summary statistics across the population. % @SPJ: I am not sure what argument this is referring to? The fact that there may or may not be a dominant lineage that differs from the mean?
%@ELD: The point was just that sometimes the dominant lineage is different from the mean and that probably matters but we're not going to talk about it. Since we're going to be short on space, I think we can just cut these lines.

@@

These metrics operate on entire phylogenies rather than single lineages within a population.
%Thus, in most cases, summarizing them across the population is not an issue as it is with lineage metrics.
Thus, there is no need to identify a representative organism or lineage.
Because they %factor in 
use date from
the entire population, phylogeny metrics can be more computationally expensive to calculate than metrics about a single lineage. On the other hand, because most lineages in the population share substantial history, phylogeny metrics can usually be calculated more rapidly than lineage metrics %can be calculated across
for an entire population. Note that phylogenies can be constructed with regard to any taxonomic level of organization, be it individual, genotype, phenotype, etc. Thus, when we refer generally to items in a phylogeny, we will use the term \textit{taxa}.

A standard technique for saving memory and time when working with phylogenties in computational systems is to ``prune" them, removing 
%any taxa that are not either currently alive or the ancestor of taxa that are alive.
% either dead (extinct) branches, or ancestors common to the entire population.
%@CAO: (NEVERMIND - SIMPLIFYING FURTHER... Does the substition I made above make sense?  I'm not quite sure what was meant by "the ancestor of taxa that are alive".
dead (extinct) branches.
Since all of the phylogeny metrics we discuss here are borrowed from natural systems (where we do not have information about taxa without offspring), they all are designed to work on pruned phylogenies. Thus, for the remainder of this paper, we will assume we are working with pruned phylogenies.

@@

An important distinction between phylogenies in natural versus computational systems to be aware of is that natural phylogenies are generally inferred from extant taxa, whereas computational phylogenies are directly recorded. The inferred phylogenies used in natural systems do not contain internal nodes except at branch points. They also do not contain history prior to the most recent common ancestor (MRCA) of all organisms in the extant population. For consistency, we will exclude pre-MRCA taxa from our analyses. However, we will not remove non-branching internal nodes, as these only serve to make our phylogenies more informative.

@@

The depth of the MRCA %(i.e. how many steps from the beginning of time did it occur?) 
(i.e., the number of steps it is from the original ancestor)
%is a useful metric because of the relatively high ratio of informativeness to ease of calculation.
is an informative metric and is easy to calculate.
A recent MRCA implies frequent selective sweeps and less long-term stable coexistence between clades. %The depth of the MRCA is easily calculated because most implementations of phylogeny pruning require keeping track of it already. 
Measuring the frequency with which the MRCA changes (i.e. the number of coalescence events) can also be informative, as some conditions can inflate the length of the lineage relative to other conditions without actually increasing the frequency of selective sweeps. This scenario is particularly likely when the population size is changing over time. A downside to the depth of MRCA as a metric is that any population that does have a stable ecology will likely never change its MRCA after the very beginning of evolution %(which is informative as a way of detecting stably coexisting branches in the phylogeny).
(which at least allows us to detect stably coexistance in the population).

@@

Measurements of phylogenetic richness quantify the total amount of evolutionary history contained in a set of taxa. The most traditional metric of phylogenetic richness is the one simply titled ``Phylogenetic Diversity", which is calculated as the number of nodes in the minimum spanning tree from the MRCA to all extant taxa~\citep{faith_conservation_1992}. Another approach is to calculate the pairwise distances between all taxa and sum them~\citep{tucker_guide_2017}. A third approach is to sum evolutionary distinctiveness, a measurement of a taxon's evolutionary uniqueness~\citep{isaac_mammals_2007}, across all extant taxa~\citep{tucker_guide_2017}.

@@

Measurements of phylogenetic divergence quantify how distinct the taxa in the population are from each other and are often %the means of values that can be calculated for
averaged across
individual taxa. For example, one option is to 
%calculate the mean of the pairwise distances between 
average the pairwise distances across
all taxa in the population~\citep{webb_exploring_2000}. Similarly, phylogenetic divergence can be calculated by 
%calculating the evolutionary distinctiveness of each taxon in the population (as alluded to in the previous section) and taking the average of these values.
% @CAO: "can be calculated by calulating" ?
averaging the evolutionary distinctiveness across each taxon in the population.
%(as alluded to in the previous section).

@@

\section{Test Problems}
\begin{figure}
\includegraphics[width=3.5in]{figs/landscapes.png}
\caption{\small The fitness landscapes used in this experiment: A) Himmelblau, B) Six-humped Camel Back, C) Shubert, and D) Composition Function 2. Interactive versions available at \url{https://emilydolson.github.io/fitness_landscape_visualizations}.}
\label{fig:landscapes}
\end{figure}

To understand the metrics defined above, the test problems used need to be well understood and studied. The benchmark functions from the GECCO Competition on Niching Methods meet both of these requirements and allow us to visualize the actual fitness landscape, due to the low dimensionality of the problems ~\citep{li_benchmark_2013}. For each problem, the X and Y coordinates offered by a given organism are translated by the function into a fitness value. We chose a diverse subset of these functions (Himmelblau, Shubert, Composition Function 2, and Six-Humped Camel Back) as our test problems in order to gain a broad understanding of our metrics. We used the implementations of these problems at \url{https://github.com/mikeagn/CEC2013} (C++ for fitness calculations during evolution, Python for post-hoc analysis). Figure \ref{fig:landscapes} shows the fitness landscapes defined by each of our four chosen test problems.

% @AML: It would be neat to have a tournament size = population size treatment (pure hill climbing)
For each test problem, we evolved populations of 1000 digital organisms under a range of mutation rates and selection strengths for 5000 generations. Each digital organism's genome consisted of two floating point numbers that defined its position in the fitness landscape. 
We initialized populations by randomly generating a number of organisms equal to the population size. 
To determine which organisms reproduced each generation, we used tournament selection. We evolved populations under five different tournament sizes: one, two, four, eight, and sixteen.  Tournament size represents strength of selection where higher tournament sizes correspond to a higher strength of selection and lower tournament sizes correspond to a lower strength of selection~\citep{blickle1995mathematical}.
A tournament size of one is equivalent to no selection pressure (\textit{i.e.} every organism in the population has an equal chance of being selected to reproduce). Organisms selected to reproduce did so asexually. Each value in an offspring's genome was mutated by adding noise given by a normal distribution with a mean of 0; the `mutation rate' of a treatment defined the standard deviation used to define this normal distribution and was given as a proportion of the test problem's domain. 
We prevented mutations from causing a value to exceed the valid domain of the given problem. 
For each problem and tournament size, we evolved populations at eight mutation rates: $1\text{e-}08$, $1\text{e-}07$, $1\text{e-}06$, $1\text{e-}05$, $1\text{e-}04$, $1\text{e-}03$, $1\text{e-}02$, and $1\text{e-}01$. 
% [.10, 0.01, 0.001, 0.0001, 0.00001,  0.000001, 0.0000001, 0.00000001]
For all experimental conditions, we ran ten replicates, each with a unique random number seed. 
Our experiment is implemented using the Empirical library; our implementation can be found at \url{https://github.com/stevenjson/ALife2018-Lineage}. % @AML: We might want to rename the repository to match whatever the title of this paper ends up being (e.g. ALIFE2018-TITLE).
